Nail: A practical interface generator for data formats

### 2014-12-03 paper/login/design.tex
pass thru design

\subsection{Redundant Data}
\label{s:dependent}
Data formats often contain values that are determined by other values or the layout of information,
such as checksums, duplicated information, or offset and  length fields. Exposing such values risks
inconsistencies that could trick the program into unsafe behavior. Therefore, we represent such values using \emph{dependent fields} and handle them transparently during
parsing and generation without exposing them to the application. 
Dependent fields are handled like other fields when parsing input, but only stored temporarily
instead of in the data type. Their value can be referenced by other parsers until it goes out of scope.
When generating output, Nail inserts the correct value.

In DNS packets, the packet header contains count fields (\cc{qc},
\cc{ac}, \cc{ns}, and \cc{ar}), which contain the number of questions and answers that follow the
header, which we represent by dependent fields (line 12-15).
Dependent fields are defined within a structure like normal fields, but their name starts with an \cc{@} symbol.
A dependent field is in scope and can be referred to by the definition of all subsequent fields in
the same structure. Dependent fields can be passed to rule invocations as parameters.

Nail allows handling count fields with \cc{n\_of}, which parses an exact number of repetitions of a rule. Lines 16-19 in Figure~\ref{fig:dns-full} shows how to use \cc{n_of} to parse the question and
answer records in a DNS packet.
Other dependencies, such as offset fields or checksums, are not handled directly by combinators, but
through  transformations, as we describe next.

@@

For a parser to be useful, it needs to support all these ways of structuring a format. This makes data description languages like
PADS~\cite{Fisher:2005:PDL:1064978.1065046} contain not just a kitchen sink, but a kitchen store
full of features -- and a language that can handle all possible formats will be a general
purpose programming language.  Instead, we keep Nail itself small, and introduce an
interface that allows complicated format structures to be handled by  plug-in \emph{transformations} 
in a general purpose
language. Of course, we ship Nail with a handy library of common transformations to handle common
format features, such as offsets,sizes and checksums.

These \emph{transformations} consume and produce
\emph{streams} -- sequences of bytes -- which can be further passed to other transformations and
eventually parsed by a Nail rule. Transformations can also access values in dependent fields. 
Streams can be subsets of other streams, for example the substream starting at an
offset given in a dependent field to handle pointer fields, or computed at runtime
, such as by decompressing another stream with zlib.

@@

As a concrete example, we will show a grammar for ProtoZIP, a very simple archive format inspired by ZIP in
Figure~\ref{fig:zip-extract}. ProtoZIP consists of a variable-length end-of-file directory, which
is a magic number followed by an array of filenames and pointers to compressed files. A grammar for the real ZIP format, which
has more layers of indirection, is presented in the full paper.

 The grammar first calls the zipdir transform on line 2, which finds the magic number and splits the file into
 two streams, one containing the compressed files, the other the directory.  Streams are referred to
 with \cc{\$identifiers}, similar to dependent fields. A C prototype of the \cc{zipdir} transform is
 shown in\ref{fig:zip-transform}. 
\begin{figure}[tb]
\smaller[0.5]
\input{code/zip-eod-xform}
\caption{Signatures of stream transform functions for handling the
end-to-beginning structure of ProtoZIP files.}
\label{fig:zip-transform}
\end{figure}
When parsing input, this will call \cc{zipdir\_parse}, which takes \cc{\$current} -- an implicit
identifier always referring to the stream currently being handled -- and returns \cc{\$files} and
\cc{\$header}. When generating output, this will call \cc{zipdir\_generate}, which appends
\cc{\$files} and \cc{\$header} to \cc{\$current}.

Line 3 then applies the \cc{dir} rule to  the \cc{\$header} stream, passing it the \cc{\$files}
stream. Within \cc{dir}, \cc{\$current} is now \cc{\$header} and input is parsed from and output
generated to that stream. The \cc{dir} rule in turn describes the structure of the directory -- a
magic number and a count field, followed by that many file descriptors. Each file descriptor, is then parsed with two
transformations, the standard-library \cc{slice} which describes an offset and a size within another
stream, and the custom \cc{zlib}, which compresses a stream using zlib. Finally, we apply a trivial
grammar (line 14) to the contents.

In a more complicated example, such as an Office document, we could now specify grammars for each
entry within an archive.

>>>>

\subsection{Redundant data}
\label{s:dependent}
Data formats often contain values that are determined by other values or the layout of information,
such as checksums, duplicated information, or offset and  length fields. Exposing such values risks
inconsistencies that could trick the program into unsafe behavior. Therefore, we represent such values using \emph{dependent fields} and handle them transparently during
parsing and generation without exposing them to the application. 
Dependent fields are handled like other fields when parsing input, but only stored temporarily
instead of in the data type. Their value can be referenced by other parsers until it goes out of scope.
When generating output, Nail inserts the correct value.

In DNS packets, the packet header contains count fields (\cc{qc},
\cc{ac}, \cc{ns}, and \cc{ar}), which contain the number of questions and answers that follow the
header, which we represent by dependent fields (lines 12--15).
Dependent fields are defined within a structure like normal fields, but their names start with an \cc{@} symbol.
A dependent field is in scope and can be referred to by the definition of all subsequent fields in
the same structure. Dependent fields can be passed to rule invocations as parameters.

Nail allows handling count fields with \cc{n\_of}, which parses an exact number of repetitions of a rule. Lines 16--19 in Figure~\ref{fig:dns-full} shows how to use \cc{n_of} to parse the question and
answer records in a DNS packet.
Other dependencies, such as offset fields or checksums, are not handled directly by combinators, but
through  transformations, as we describe next.

@@

For a parser to be useful, it needs to support all these ways of structuring a format. This makes data description languages like
PADS~\cite{Fisher:2005:PDL:1064978.1065046} contain not just a kitchen sink, but a kitchen store
full of features, and a language that can handle all possible formats will be a general-purpose
programming language.  Instead, we keep Nail itself small, and introduce an
interface that allows complicated format structures to be handled by  plug-in \emph{transformations} 
in a general-purpose
language. Of course, we ship Nail with a handy library of common transformations to handle common
format features, such as offsets,sizes and checksums.

These \emph{transformations} consume and produce
\emph{streams}---sequences of bytes---which can be further passed to other transformations and
eventually parsed by a Nail rule. Transformations can also access values in dependent fields. 
Streams can be subsets of other streams, for example the substream starting at an
offset given in a dependent field to handle pointer fields, or computed at runtime,
such as by decompressing another stream with zlib.

@@

As a concrete example, we will show a grammar for ProtoZIP, a very simple archive format inspired by ZIP in
Figure~\ref{fig:zip-extract}. ProtoZIP consists of a variable-length end-of-file directory, which
is a magic number followed by an array of filenames and pointers to compressed files. A grammar for the real ZIP format, which
has more layers of indirection, is presented in the full paper~\cite{bangert:nail}.

The grammar first calls the \cc{zipdir} transform on line 2, which finds the magic number and splits the file into
two streams, one containing the compressed files, the other the directory.  Streams are referred to
with \cc{\$identifiers}, similar to dependent fields. A C prototype of the \cc{zipdir} transform is
shown in Figure~\ref{fig:zip-transform}.

\begin{figure}[tb]
\smaller[0.5]
\input{code/zip-eod-xform}
\caption{Signatures of stream transform functions for handling the
end-to-beginning structure of ProtoZIP files.}
\label{fig:zip-transform}
\end{figure}
When parsing input, line 2 of Figure~\ref{fig:zip-extract} will call \cc{zipdir\_parse}, which takes \cc{\$current} (an implicit
identifier always referring to the stream currently being handled) and returns \cc{\$files} and
\cc{\$header}. When generating output, the same line will call \cc{zipdir\_generate}, which appends
\cc{\$files} and \cc{\$header} to \cc{\$current}.

Line 3 of Figure~\ref{fig:zip-extract} then applies the \cc{dir} rule to  the \cc{\$header} stream, passing it the \cc{\$files}
stream as an argument. Within \cc{dir}, \cc{\$current} is now \cc{\$header} and input is parsed from and output
generated to that stream. The \cc{dir} rule in turn describes the structure of the directory, which consists of a
magic number and a count field, followed by that many file descriptors. Each file descriptor is then parsed with two
transformations, the standard-library \cc{slice} which describes an offset and a size within another
stream, and the custom \cc{zlib}, which compresses a stream using zlib. Finally, we apply a trivial
grammar (line 14) to the contents.

In a more complicated example, such as an Office document, we could now specify grammars for each
file contained within an archive.
